Hybrid recurrent model for regime switching nonstationary time series

### 2021-09-26 iclr2022/hybridmodel_iclr.tex
Add infos on the classifier RNN

\newcommand{\ts}{y}
\newcommand{\fullts}{{\bf \ts}}
\newcommand{\tspred}{\hat{\ts}}
\newcommand{\lvl}{\hat{l}}
\newcommand{\fulllvl}{{\bf \lvl}}
\newcommand{\seas}{\hat{s}}
\newcommand{\fullseas}{{\bf \seas}}
\newcommand{\lsmooth}{\alpha}
\newcommand{\ssmooth}{\beta}
\newcommand{\lag}{h}
\newcommand{\window}{w}
\newcommand{\etspred}{\hat{y}}
\newcommand{\fulletspred}{{\bf \etspred}}
\newcommand{\tswindow}{{\bf \ts}}
\newcommand{\meants}{\Bar{\ts}}
\newcommand{\rnnwindow}{{\bf \rnn}}
\newcommand{\rnn}{z}
\newcommand{\err}{err}
\newcommand{\errwindow}{{\bf \err}}
\newcommand{\rnnmodel}{\textsc{rnn}}
\newcommand{\etsmodel}{\textsc{ets(n,n,a)}}
\newcommand{\ws}{w}
\newcommand{\fullws}{{\bf \ws}}
\newcommand{\wswindow}{{\bf \ws}}
\newcommand{\concatinput}{x}
\newcommand{\fullconcatinput}{{ \bf \concatinput}}
\newcommand{\numberts}{14000}

@@

\subsection{Per time series statistical model}
%The first step in our additive hybrid model is to consider, for each time series, an exponential smoothing  statistical model.
The time-series-specific models compute, for each sequence, two components,  level  and  seasonality. For all $1\leqslant n \leqslant N$ and $0\leqslant t \leqslant T$, let $\lvl_t^n$ be the value of the $n$-th level component at time $t$ and  $\fulllvl^n = \{\lvl_t^n\}_{0\leqslant t \leqslant T}$ be all the values of this level component. For all $1\leqslant n \leqslant N$ and $-m+1\leqslant t \leqslant T$, let $\seas_t^n$ be the value of the $n$-th seasonality component at time $t$ and  $\fullseas^n = \{\seas_t^n\}_{-m+1\leqslant t \leqslant T}$ be all the values of this season component. $m \in \mathbb{N}$ is the seasonality (4 for quarterly, 12 for monthly or 52 for weekly time series). Given the sequence $\fullts^n$, the two components  $\fulllvl^n$ and $\fullseas^n$ are computed using exponential smoothing.
For all $1\leqslant n \leqslant N$, let $\Theta^n = \{\alpha^n, \gamma^n, \lvl^n_0, \{\seas_t^n\}_{-m+1\leqslant t \leqslant 0}\}$ be the parameters of the $n$-th sequence with $\lsmooth^n$ and $\ssmooth^n$ the positive smoothing parameters, $\lvl^n_0$ the first level term and $\{\seas_t^n\}_{-m+1\leqslant t \leqslant 0}$ the $m$ first seasonality terms. We have finally for all $1\leqslant n \leqslant N$ and $1\leqslant t \leqslant T$, \textcolor{blue}{Now, I initialize alpha, gamma, the 52 first seasons and the first level with a python package}
%We initialize the first level with $\lvl^n_{0} = \sum_{t = 1}^m \ts^n_{t}/m$ \textcolor{red}{issue with the time index ?} and, for all $1\leqslant n \leqslant N$ and $1\leqslant t \leqslant T$,
\begin{align*}
    \lvl^n_{t} &= \lsmooth^n (\ts^n_t - \seas^n_{t-m}) + (1- \lsmooth^n)\lvl^n_{t-1}\,, \\
    \seas^n_t &= \ssmooth^n (\ts^n_t - \lvl^n_{t-1}) + (1 - \ssmooth^n)\seas^n_{t-1}\,. 
\end{align*}

@@

Let $\window \in \mathbb{N}$ be the window size, usually this input window is equal to the forecast horizon $\window = \lag$. Consider the following  normalized, deseasonalized and rescaled sequence $\rnnwindow^n_T = \{\rnn^{n}_{T-\window+i|T}\}_{1\leqslant i \leqslant w}$, where, for all $1\leqslant n \leqslant N$ and $1\leqslant i \leqslant w$, 
$$
\rnn^{n,T}_{T-w+i|T} = \frac{\ts^n_{T-w+i} - \lvl^n_T - \seas^n_{T-w+i}}{\meants^n_T}\,,
$$
where $ \meants^n_T = \sum_{i = 1}^{w}  \ts^n_{T-w+i}/w$. This new input is used to correct the exponential smoothing prediction. %Write, for all $1\leqslant n \leqslant N$ and all $i \in \{1,\ldots,H\}$, $\err^{n,T}_{T+i|T} = \ts^n_{T+i} - \etspred^{ets,n}_{T+i|T}$. 
Consider then a recurrent neural network denoted by RNN-p which aims at predicting the exponential smoothing error. Given $\rnnwindow^n_T$, the proposed model is given, for all $1\leqslant n \leqslant N$ and all $i \in \{1,\ldots,\lag\}$, by
$$
\ts^n_{T+i}  = \etspred^{ets,n}_{T+i|T} +  U^n_T\times\meants^n_T\times \rnnmodel(\rnnwindow^n_T)_{T+i} + \varepsilon^n_{T+i}\,,%\\
%  &= \lvl^n_T + \seas^n_{T+i-m(k+1)} +  U^n_T\times\meants^n_T\times \rnnmodel(\rnnwindow^n_T)_{T+i} + \varepsilon^n_{T+i}\,,
$$
where $\{\varepsilon^n_{T+i}\}_{i\leqslant 1}$ are independent random variables with a Laplace distribution and $U^n_T$ an unobserved regime switching Bernoulli random variable. The random variable $U^n_T$ allows to provide RNN-p corrections only when ETS prediction is not sharp enough \textcolor{red}{Show in appendix cases where ETS alone is enough and cases where correction is required}. The law of $U^n_T$ is parameterized by a recurrent neural network denoted by RNN-c with the same input than RNN-p \textcolor{blue}{RNN-p for the predictor and RNN-c for the classifier, maybe names are confusing. They have, for now, the same archi/inputs}.
%$$
%\ts^n_{T+i} = \rnnmodel(\rnnwindow^n_T)_{T+i} = \frac{\hat{\err}^{n,T}_{T+i|T}}{\meants^n_T}
%$$
%Finally, for all $1\leqslant n \leqslant N$, the  final forecast of the model is $\{\tspred^{n}_{T+1:T+\lag|T}\}_{1\leqslant n \leqslant N}$, where, for all $i \in \{1,\ldots,\lag\}$ :
%\begin{align*}
%\tspred^n_{T+i|T}  &= \etspred^{ets,n}_{T+i|T} +  \meants^n_T\times \rnnmodel(\rnnwindow^n_T)_{T+i}\,,\\
%  &= \lvl^n_T + \seas^n_{T+i-m(k+1)} + \hat{\err}^{n,T}_{T+i|T}\,.
%\end{align*}

@@

Using a time-series-specific statistical model, the hybrid network achieves striking accuracy on single sequences. Then, the shared recurrent neural network relies on the strength of machine learning methods at leveraging a huge quantities of heterogeneous data.
In addition to the $N$ target time series, we consider $K \times N$ external sequences indexed from $0$ to $T$. For all $1\leqslant n \leqslant N$, $1\leqslant k \leqslant K$ and  $1\leqslant t \leqslant T$, let $\ws^{n,k}_t$ be the value of the $k$-th external sequence at time $t$ linked to the sequence $\fullts^n$. We note $\fullws^n = \{\{\ws_t^{n,k}\}_{1\leqslant t \leqslant T}\}_{1\leqslant k \leqslant K}$ all the values of the family of weak signals linked to the $n$-th sequence. In addition, let $\fullws^n_T = \{\{\ws_{T-w+i}^{n,k}\}_{1\leqslant i \leqslant \window}\}_{1\leqslant k \leqslant K}$ be only the last $\window$ terms of the family. Concatenating $ \rnnwindow^n_T$ and $\fullws^n_T$, we define a new input for our RNN,   
\begin{align*}
\fullconcatinput^n_T &= \{\concatinput^n_{T-w+i|T}\}_{1\leqslant i \leqslant w}\,, \\
&= \{\rnn^n_{T-w+i|T}, \ws^{n,1}_{T-w+i},...,\ws^{n,K}_{T-w+i}\}_{1\leqslant i \leqslant w}\,.
\end{align*}
Finally, for all $1\leqslant n \leqslant N$ and for all $i \in \{1,\ldots,\lag\}$ our final model becomes:
%$$
%\tspred^n_{T+i|T} 
%= \etspred^{ets,n}_{T+i|T} +  \rnnmodel(\fullconcatinput^n_T)_{T+i} \times \meants^n_T
%$$
\begin{equation}
\label{eq:full:model}
\ts^n_{T+i}  = \etspred^{ets,n}_{T+i|T} +  U^n_T\times\meants^n_T\times \rnnmodel(\fullconcatinput^n_T)_{T+i} + \varepsilon^n_{T+i}\,,
\end{equation}
\textcolor{red}{specify here the right distribution for $U$ (depends on the weak signal or no)}. The complete architecture is displayed in Figure~\ref{fig:architecture}.

>>>>

\newcommand{\ts}{y}
\newcommand{\fullts}{{\bf \ts}}
\newcommand{\tspred}{\hat{\ts}}
\newcommand{\lvl}{\hat{l}}
\newcommand{\fulllvl}{{\bf \lvl}}
\newcommand{\seas}{\hat{s}}
\newcommand{\fullseas}{{\bf \seas}}
\newcommand{\lsmooth}{\alpha}
\newcommand{\ssmooth}{\beta}
\newcommand{\lag}{h}
\newcommand{\window}{w}
\newcommand{\etspred}{\hat{y}}
\newcommand{\fulletspred}{{\bf \etspred}}
\newcommand{\tswindow}{{\bf \ts}}
\newcommand{\meants}{\Bar{\ts}}
\newcommand{\rnnwindow}{{\bf \rnn}}
\newcommand{\rnn}{z}
\newcommand{\err}{err}
\newcommand{\errwindow}{{\bf \err}}
\newcommand{\rnnmodel}{\textsc{rnn}}
\newcommand{\etsmodel}{\textsc{ets(n,n,a)}}
\newcommand{\ws}{w}
\newcommand{\fullws}{{\bf \ws}}
\newcommand{\wswindow}{{\bf \ws}}
\newcommand{\concatinput}{x}
\newcommand{\fullconcatinput}{{ \bf \concatinput}}
\newcommand{\numberts}{14000}
\newcommand{\threshold}{\eta}

@@

\subsection{Per time series statistical model}
%The first step in our additive hybrid model is to consider, for each time series, an exponential smoothing  statistical model.
The time-series-specific models compute, for each sequence, two components,  level  and  seasonality. For all $1\leqslant n \leqslant N$ and $0\leqslant t \leqslant T$, let $\lvl_t^n$ be the value of the $n$-th level component at time $t$ and  $\fulllvl^n = \{\lvl_t^n\}_{0\leqslant t \leqslant T}$ be all the values of this level component. For all $1\leqslant n \leqslant N$ and $-m+1\leqslant t \leqslant T$, let $\seas_t^n$ be the value of the $n$-th seasonality component at time $t$ and  $\fullseas^n = \{\seas_t^n\}_{-m+1\leqslant t \leqslant T}$ be all the values of this season component. $m \in \mathbb{N}$ is the seasonality (4 for quarterly, 12 for monthly or 52 for weekly time series).
For all $1\leqslant n \leqslant N$, let $\Theta^n = \{\alpha^n, \gamma^n, \lvl^n_0, \{\seas_t^n\}_{-m+1\leqslant t \leqslant 0}\}$ be the parameters of the $n$-th sequence with $\lsmooth^n$ and $\ssmooth^n$ the positive smoothing parameters, $\lvl^n_0$ the first level term and $\{\seas_t^n\}_{-m+1\leqslant t \leqslant 0}$ the $m$ first seasonality terms. Given the sequence $\fullts^n$ and the parameters $\Theta^n$ , the two components  $\fulllvl^n$ and $\fullseas^n$ are computed as follow :
%We initialize the first level with $\lvl^n_{0} = \sum_{t = 1}^m \ts^n_{t}/m$ \textcolor{red}{issue with the time index ?} and, for all $1\leqslant n \leqslant N$ and $1\leqslant t \leqslant T$,
\begin{align*}
    \lvl^n_{t} &= \lsmooth^n (\ts^n_t - \seas^n_{t-m}) + (1- \lsmooth^n)\lvl^n_{t-1}\,, \\
    \seas^n_t &= \ssmooth^n (\ts^n_t - \lvl^n_{t-1}) + (1 - \ssmooth^n)\seas^n_{t-1}\,. 
\end{align*}

@@

Let $\window \in \mathbb{N}$ be the window size, usually this input window is equal to the forecast horizon $\window = \lag$. Consider the following  normalized, deseasonalized and rescaled sequence $\rnnwindow^n_T = \{\rnn^{n}_{T-\window+i|T}\}_{1\leqslant i \leqslant w}$, where, for all $1\leqslant n \leqslant N$ and $1\leqslant i \leqslant w$, 
$$
\rnn^{n,T}_{T-w+i|T} = \frac{\ts^n_{T-w+i} - \lvl^n_T - \seas^n_{T-w+i}}{\meants^n_T}\,,
$$
where $ \meants^n_T = \sum_{i = 1}^{w}  \ts^n_{T-w+i}/w$. This new input is used to correct the exponential smoothing prediction. %Write, for all $1\leqslant n \leqslant N$ and all $i \in \{1,\ldots,H\}$, $\err^{n,T}_{T+i|T} = \ts^n_{T+i} - \etspred^{ets,n}_{T+i|T}$. 
Consider then a recurrent neural network denoted by RNN-p which aims at predicting the exponential smoothing error. Given $\rnnwindow^n_T$, the proposed model is given, for all $1\leqslant n \leqslant N$ and all $i \in \{1,\ldots,\lag\}$, by
$$
\ts^n_{T+i}  = \etspred^{ets,n}_{T+i|T} +  U^n_T\times\meants^n_T\times \rnnmodel_p(\rnnwindow^n_T)_{T+i} + \varepsilon^n_{T+i}\,,%\\
%  &= \lvl^n_T + \seas^n_{T+i-m(k+1)} +  U^n_T\times\meants^n_T\times \rnnmodel_p(\rnnwindow^n_T)_{T+i} + \varepsilon^n_{T+i}\,,
$$
where $\{\varepsilon^n_{T+i}\}_{i\leqslant 1}$ are independent random variables with a Laplace distribution and $U^n_T$ an unobserved regime switching Bernoulli random variable. The random variable $U^n_T$ allows to provide RNN-p corrections only when ETS prediction is not sharp enough \textcolor{red}{Show in appendix cases where ETS alone is enough and cases where correction is required}. The law of $U^n_T$ is parameterized by a recurrent neural network denoted by RNN-c with the same input than RNN-p.
\begin{align*}
U^n_T = \mathbf{1}_{\rnnmodel_p(\rnnwindow^n_T) > \threshold}
\end{align*}

\textcolor{blue}{RNN-p for the predictor and RNN-c for the classifier, maybe names are confusing. They have, for now, the same archi/inputs}.
%$$
%\ts^n_{T+i} = \rnnmodel(\rnnwindow^n_T)_{T+i} = \frac{\hat{\err}^{n,T}_{T+i|T}}{\meants^n_T}
%$$
%Finally, for all $1\leqslant n \leqslant N$, the  final forecast of the model is $\{\tspred^{n}_{T+1:T+\lag|T}\}_{1\leqslant n \leqslant N}$, where, for all $i \in \{1,\ldots,\lag\}$ :
%\begin{align*}
%\tspred^n_{T+i|T}  &= \etspred^{ets,n}_{T+i|T} +  \meants^n_T\times \rnnmodel(\rnnwindow^n_T)_{T+i}\,,\\
%  &= \lvl^n_T + \seas^n_{T+i-m(k+1)} + \hat{\err}^{n,T}_{T+i|T}\,.
%\end{align*}

@@

Using a time-series-specific statistical model, the hybrid network achieves striking accuracy on single sequences. Then, the shared recurrent neural network relies on the strength of machine learning methods at leveraging a huge quantities of heterogeneous data.
In addition to the $N$ target time series, we consider $K \times N$ external sequences indexed from $0$ to $T$. For all $1\leqslant n \leqslant N$, $1\leqslant k \leqslant K$ and  $1\leqslant t \leqslant T$, let $\ws^{n,k}_t$ be the value of the $k$-th external sequence at time $t$ linked to the sequence $\fullts^n$. We note $\fullws^n = \{\{\ws_t^{n,k}\}_{1\leqslant t \leqslant T}\}_{1\leqslant k \leqslant K}$ all the values of the family of weak signals linked to the $n$-th sequence. In addition, let $\fullws^n_T = \{\{\ws_{T-w+i}^{n,k}\}_{1\leqslant i \leqslant \window}\}_{1\leqslant k \leqslant K}$ be only the last $\window$ terms of the family. Concatenating $ \rnnwindow^n_T$ and $\fullws^n_T$, we define a new input for our RNN,   
\begin{align*}
\fullconcatinput^n_T &= \{\concatinput^n_{T-w+i|T}\}_{1\leqslant i \leqslant w}\,, \\
&= \{\rnn^n_{T-w+i|T}, \ws^{n,1}_{T-w+i},...,\ws^{n,K}_{T-w+i}\}_{1\leqslant i \leqslant w}\,.
\end{align*}
Finally, for all $1\leqslant n \leqslant N$ and for all $i \in \{1,\ldots,\lag\}$ our final model becomes:
%$$
%\tspred^n_{T+i|T} 
%= \etspred^{ets,n}_{T+i|T} +  \rnnmodel(\fullconcatinput^n_T)_{T+i} \times \meants^n_T
%$$
\begin{equation}
\label{eq:full:model}
\ts^n_{T+i}  = \etspred^{ets,n}_{T+i|T} +  U^n_T\times\meants^n_T\times \rnnmodel_p(\fullconcatinput^n_T)_{T+i} + \varepsilon^n_{T+i}\,,
\end{equation}
with :
\begin{align*}
U^n_T = \mathbf{1}_{\rnnmodel_p(\fullconcatinput^n_T) > \threshold}
\end{align*}
The complete architecture is displayed in Figure~\ref{fig:architecture}.
